Field and experimental data together with computational models reveal how cleaner fish adjust decisions in a biological market

### 2022-03-01 manuscript_1.0/manuscript_1.0.tex
Edited the manuscript according to Red1 comments. Readme.md file to facilitate reproducibility.

%%%% Article title to be placed here
\title{Combination of field and experimental data together with computational models reveal cognitive mechanisms behind cleaner fish behaviour}

@@

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings yield similar performances, and Lloyd Morgan's cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish \emph{Labroides dimidiatus}, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner fish relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Often alternative cognitive mechanisms yield similar behaviour and/or cognitive
performances. This poses a problem for disentangling the mechanistic
underpinnings of behaviour. This is particularly clear in research aimed at
discovering interspecific variation in \emph{higher} cognitive abilities; or
in other words, research on whether non-human animals show cognitive
abilities believed to be uniquely human. For instance, when researchers
try to find \emph{mental time travel-like} behaviour, they usually come up with
clever experiments showing the behaviour displayed requires inferences made
through past events\citep{dally_Foodcaching_2006}. However, they often face challenges proposing alternative scenarios where simple associations can bring about the
predicted behavioural outcome \citep{suddendorf_Evolution_2007}. Similarly, attempts
to show \emph{theory of mind} in non-human animals face objections
justified by alternative mechanisms that bring about similar behavioural
results \citep{heyes_Theory_1998}. Such controversies are usually settled by
using the principle of parsimony, and it's cognitive version,
Lloyd Morgan's cannon. Which states that the simpler explanation (mechanism)
must be accepted. Ideally, however, alternative hypothesis should be evaluated
in light of their explanatory power. One potential dimension that could allow
the variation necessary among distinct cognitive mechanisms is environmental
variation. Controlled laboratory conditions often mask the inter and
intra-specific variation that arises in the interaction between mechanisms and
environments. Thus, if predictions from the alternative mechanisms change
with environmental conditions, it is possible to asses the explanatory power
of the different hypothesis by performing behavioural test in the context of
different environmental conditions.

From a evolutionary perspective, mechanisms are likely selected because
of how they allow individuals to respond to environmental variation.
For example, biological market theory predicts that the exchange rate of goods
and/or services traded between cooperative partners adjusts to the
law of supply and demand, when individuals have some degree of partner choice
\citep{noe_Biological_1995}. Supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should favour the ability
to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent
mechanisms beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Client fish seek cleaner fish services at their territory
(so-called ``cleaning station'') and offer themselves as food patches
to get their ectoparasites removed, which provides cleaners
with food and clients with improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016, demaire_Reduced_2020}.
Given the capacity of some client fish to swim larger distances and
access multiple cleaning stations while others access the only cleaning
station at their territory, it is crucial to categorize clients as
either ``visitors'' or ``residents'', respectively. During cleaning interactions,
a cleaner fish often face a choice between a visitor and a resident client
seeking its cleaning services simultaneously. Visitors have the option
to switch to another cleaner fish if being made to wait, while residents
must wait for inspection. Indeed,
visitors have been observed to use their partner choice option in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001}.

When researchers aimed at testing wild-caught cleaner fish, as
well as individuals from other species, abilities
to prioritize visitor over resident clients in a lab-based paradigm,
they used Plexiglas plates of different colours and shapes offering same
amount of food as surrogates for visitor and resident clients.
One plate is made to behave like a resident, i.e.~it remains until
the cleaner fish had eaten off it. The other plate was made to behave
like a visitor, i.e.~it is removed if not inspected first.
This laboratory paradigm originally inspired by the cleaning market
of cleaner fish is more generally refered as the ephemeral reward task.
Individuals are assesed in their capacity to develop a preference for the
ephemeral option, which gives them access to two reward units.
Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Furthermore, adult cleaner fish outperformed various primates as well as rats and
pigeons in this original version of the ephemeral reward task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; yet the African grey
parrots solve this task as well \citep{pepperberg_Can_2014}. This interspecific
variation in performance, and primates' known cognitive abilities \(\color{red}{\text{ref}}\), shows that performane in the task
cannot be readily explained by classifying species as having high and
low cognitive abilities. Instead, this inter-spefic variation suggest
that solving the ephemeral reward task requires a set of cognitive
mechanisms that are not present in all the tested taxa, or were not used
under the experimental set-up.

Recent research on the cognitive tool kit needed to solve the ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009}, who defined cognition as including
all ways in which animals acquire information
through the senses, process, retain and decide to act on it.
Indeed, perception of relevant cues is of major importance
for individual performance. For example, some animals find relevant
information in the plates \citep{wismer_Cuebased_2019} while others find
it in the food \citep{pretot_Comparative_2021, pretot_Comparing_2016}.
However, identifying cues as salient is not the only challenge
of the ephemeral reward task, as revealed by proximate learning models.
Such models allow varying the cognitive tool kit and to evaluate
which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the ephemeral reward task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaner fish not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well a resident alone or
a visitor alone. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible scenarios (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to distinguish and value differently
one stimulus alone from compound versions of it has been termed
configurational learning, or chunking,
or segmentation (see references in \citep{prat_Modelling_2022}).

To solve the ephemeral reward task, cleaners need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Low temporal discounting is when
individuals include in their valuation of an action the reward effects
that this will have in the future. This is done by combining in a single
valuation the reward obtained in the current time with all the reward that
comes after, discounting for how far in the future reward
is accrued. `Chaining' the reward of these different time steps allows
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations \citep{enquist_Power_2016}. Even though, `chaining'
can be readily implemented computationally in learning models
\citep{enquist_Power_2016, sutton_Reinforcement_2018}, cognitively it seems to be
a complex adaptation \citep{suddendorf_Evolution_2007}. On the other hand,
using client behaviour as a negative reinforcer is in principle
easier to implement. Negative reinforcement is part of the ubiquitous
learning mechanism termed operant conditioning \citep{thorndike_Animal_1898, skinner_Behavior_1938}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Ideally, however, the two
mechanisms should be evaluated in light of how well
they explain the available data. Note that different fields interested in
cognition and desicion making use different words to refer to negative
reinforcers \citep{quinones_Reinforcement_2019, sutton_Reinforcement_2018}.
In here, for the sake of simplicity and clarity, we will use the word
`penalty' to refer to this mechanism which includes a negative reinforcer.

Over the last decade, our team tested over a hundred of wild-caught
cleaner fish in the exact same paradigm of of the ephemeral reward
task \citep{salwiczek_Adult_2012, wismer_Variation_2014, triki_Decrease_2018, triki_Biological_2019, triki_Brain_2020}. These fish often come from
different reef locations around the study field site at Lizard Island,
Great Barrier Reef in Australia. Further investigation of the local
eco-sociological conditions revealed that cleaner and client
fish population densities have a substantial impact on cleaner fish
performance in the task. Cleaner fish from reef sites with relatively
low densities were more likely to fail at solving the task \citep{triki_Biological_2019, triki_Decrease_2018, wismer_Variation_2014}. The explanation for this pattern
is twofold. First, low cleaner fish population densities imply
low supply and high demand for cleaning services.
Under such conditions, there are fewer occasions under
which failing to give priority to a visitor translates to an empty
cleaning station. Second and also related to the state of the cleaning market,
visitor clients at low cleaner fish population densities
are less likely to exert their partner choice options
and are hence more likely to wait for inspection if not given cleaning
priority \citep{triki_Brain_2020}. In the absence of partner choice
cleaners should not give priority to visitors any more.

The interplay between the underling cognitive machinery and
local ecological conditions apparently generate the documented variable
performance among individuals of the same species.
Our approach of fitting the computational model to the empirical data on
fish densities and cleaner fish censuses and performance
in the ephemeral reward task aimed at: (i), determining which
mechanism cleaner fish use to incorporate future consequences of current
decisions by testing whether chaining, penalty, or a combination
of both best explains their performance; (ii) determining whether the
two mechanisms differed with respect to the ecological conditions
that are likely to cause high versus low performance in the
ephemeral reward task. Additionally, we assessed which mechanism
yields optimal performance patterns. Relying on the logic of
biological market theory, we predicted
that appropriate performance is to show low preference for visitors under low
local cleaner-to-client ratio.

@@

The model consists of a set individual-based simulations where
individuals, representing cleaners, face a series of choices
between two options, which simulate the natural conditions
of the cleaning market. Individuals experience a series
of discrete time points in which they face different `states',
defined by the number and category of client fish (visitor or residents)
inviting for cleaning services. There are six possible states:
zero clients, one resident, one visitor,
resident-resident, visitor-visitor, and resident-visitor.
The probability of each state is largely determined by the relative
abundance of cleaner fish, residents and visitors, but to some degree by
cleaner fish choices when it faces the resident-visitor combination.
This is because residents are willing to queue for cleaning service;
while visitors leave the queue (with a certain probability) when made to wait.
Individuals obtain a fixed reward from cleaning a client fish
regardless of the category. Every time individuals face and make a choice
they update the probability of making that same choice. The update is
based on the difference between the expected reward and the obtained
reward (prediction error) \citep{sutton_Reinforcement_2018, rescorla_Theory_1972}. Given the new information, the update is carried
in the direction that leads to more reward being obtained, given
the new information. In the long run, the probability of choosing a visitor
over a resident converges in the model. To which probability the model
will converge depends on the relative abundance of cleaners, visitors
and resident; as well as on the probability of visitors leaving the
cleaning station when made to wait. Further details of the model
implementation can be found in Quiñones \emph{et al} \citep{quinones_Reinforcement_2019}.

The model shows that agents need to find a mean
to incorporate future consequences of current choices. In the model,
this could be achieved with either of two parameters that could
also work together. First, \(\gamma\) measures how
much individuals include future rewards in their decision updates.
If \(\gamma=0\), individuals only use the immediate reward obtained from a
cleaning interaction. As \(\gamma\) increases individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards of an action (chaining).
Second, \(\eta\) measures how much individuals include in their reward the
fleeing behaviour of visitor as a negative component (penalty). Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

@@

The cognitive performance data was from a total of 120 cleaners
(10 individuals per 12 site/year) tested in the
ephemeral reward task \citep{triki_Biological_2019, triki_Brain_2020}. Authors
housed all captured cleaners individually in glass aquaria
( \(62cm \times 27cm \times 37 cm\) ) and provided them
with PVC pipes (\(10 cm \times 1 cm\)) as shelters.
The task consisted of exposing the cleaner fish to substitute
models of client fish in the form of two \emph{Plexiglas} plates offering the
same amount of food (one item of mashed prawn). The two plates differed
in colour and pattern (horizontal green stripes or vertical pink stripes)
but had equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or
resident (permanent food source). That is, if a cleaner fish inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor plate
based solely on the behaviour-cue of the plates rather than size-cue
\citep{wismer_Cuebased_2019}. Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
tested the fish for a maximum of 200 trials with 20 trials a day, 10 trials
in the morning and 10 trials in the afternoon. They randomized and
counterbalanced the plates' spatial location (i.e.~left or right)
between trials. Similarly, they counterbalanced the plates' decoration
(colour and pattern) and the plates' role (visitor or resident) between the
tested fish. In the original studies, once a fish reach a learning criterion,
that is, performing significantly above chance level (\(> 50%
\), \(p-value ≥ 0.05\)),
they passed to a reversal version of the task where the roles of the
visitor/resident Plexiglas plates were swapped (see Triki et al.~2019, 2020).
Here, we used instead a subset of these data in order to have an idea on
cleaner fish preferences for the visitor plate, even if they do not reach
the learning criterion within 200 trials. To do so, we first extracted the
trial-by-trial outcomes from the last two session (20 trials) of those who
never reached the learning criterion for visitor plate
(\(N = 45\) cleaner fish).
For those who reached the learning criterion at some point during the
test and passed to a reversal phase, we extracted the trials-by-trial
outcomes from the last session (10 trials) before passing to reversal
and the last 10 trials they were exposed to in the test
(\(N = 75\) cleaner fish).
We chose a combination of initial and reversal to quantify preference
for the visitor client, given that it matches well the criteria chosen
in previous analysis of the ephemeral reward task experimental set-up
\citep{triki_Biological_2019, triki_Brain_2020}. In the
suplementaty material (Fig. \ref{fig:rawdata}) we show how our choice maps
to the previously used criteria. For comparison, we also show a quantification
of visitor preference based only on the initial phase of the experiments.

@@

The aim of the analyses is to fit the key model parameters
\(\gamma\) and \(\eta\), to the empirical data from Triki \emph{et al.}
\citep{triki_Biological_2019, triki_Brain_2020} to test whether each or
a combination of these effects is a better explanation for the pattern
seen in the data. We used the ecological variables: cleaners, visitor clients,
resident clients abundances and visitor clients leaving probability,\\
as input to the models. As response variable we used cleaners'
preference for visitor over residents in the ephemeral reward task.
Finally, we used the preference for visitors resulting from the
model simulations as the prediction for the response variable.
We kept all other parameter values used for the model simulations constant,
see Table \ref{tab:param}.

To capture with the model the relationship between the ecological
variables and cleaner fish preferences for visitors we need to scale
the absolute population densities of cleaner fish from the empirical data
to a measure of relative abundances that captures client visitation patterns.
This is because in the model, relative abundances of clients
define not only the probability of residents and visitors,
but also how often the cleaning station is empty (e.g.~there are no
clients to be cleaned). The frequency with which clients visit the station
is another variable, which in nature may vary among different client species
depending on their ectoparasite loads. We do not have field estimates
for species-specific parasite loads, especially not as a function of site.
In order to control for these aspects, we computed a measure of relative
cleaner fish abundance for each reef site relative to
absolute abundances and multiplied it
by a scaling constant that changes the range of the variable. This scaling
constant is hence meant to capture variation in the market conditions driven
partly by cleaner fish abundance. We fitted the value of the scaling constant as
part of the statistical inference. As for the visitor and resident
abundances we computed a relative measure with respect to the
total client abundance, and weighted that by the rescaled relative absence
of cleaners (\(1-\text{relative cleaner abundance}\)). Thus, all three measures
of relative abundance sum up to 1, and can be used in the model as a proxy
for the probability of having different options in the cleaning station.
Note that we introduced the scaling constant to control for variation that
is not captured by the model; its parameter distribution does not offer
biological insights.

>>>>

%%%% Article title to be placed here
\title{Field and experimental data together with computational models reveal how cleaner fish adjust decisions in a biological market}

@@

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings yield similar performances, and Lloyd Morgan's cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. We illustrate this point by fitting computational models to experimental data on performance by wild-caught cleaner fish Labroides dimidiatus in an ephemeral reward task, as well as cleaner and client fish densities from the locations of capture. Using Bayesian statistics to fit the model parameters to performance data revealed that cleaner fish most likely estimate future consequences of an action, while it appears unlikely that the removal of the ephemeral reward acts as psychological punishment (negative reinforcement). Incorporating future consequences also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement mechanism. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinnings of cognitive performance.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Often alternative cognitive mechanisms yield similar behaviour and/or cognitive
performances. This poses a problem for disentangling the mechanistic
underpinnings of behaviour. This is particularly clear in research aimed at
discovering between species variation in \emph{higher} cognitive abilities; or
in other words, research on whether non-human animals show cognitive
abilities believed to be uniquely human. For instance, when researchers
try to find \emph{mental time travel-like} behaviour, they usually come up with
clever experiments to show the behaviour displayed requires inferences made
through past events\citep{dally_Foodcaching_2006}. However, they often face the
challenge of alternative scenarios where simpler explanations,
like classic associative learning, can bring about the
observed behavioural outcome \citep{suddendorf_Evolution_2007}. Similarly, attempts
to demonstrate the presence of \emph{theory of mind} in non-human animals
face objections justified by alternative mechanisms underpinning
similar behavioural results \citep{heyes_Theory_1998}.
Such controversies are usually settled by using the principle
of parsimony and its cognitive version,
Lloyd Morgan's cannon, which states that the simpler explanation (mechanism)
should be accepted. Ideally, however, alternative hypotheses should be evaluated
in light of their explanatory power. One potential dimension that could allow
the variation necessary among distinct cognitive mechanisms is environmental
variation. Controlled laboratory conditions often mask between and within
species variation that arises among different environments. Thus, if
predictions from the alternative mechanisms depend on environmental
conditions, it is possible to assess their relative explanatory power
by performing cognitive test in the context of
different environmental conditions.

From an evolutionary perspective, mechanisms are likely selected because
of how they allow individuals to respond to environmental variation.
For example, biological market theory predicts that the exchange rate of goods
and/or services traded between cooperative partners adjusts to the
law of supply and demand, when individuals have some degree of partner choice
\citep{noe_Biological_1995}. Supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should favour the ability
to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying conditions. More specifically the question arises
to what extent mechanisms beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Client fish seek cleaner fish services at their territory
(so-called ``cleaning station'') and offer themselves as food patches
to get their ectoparasites removed, which provides cleaners
with food and clients with improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016, demaire_Reduced_2020}.
Given the capacity of some client fish to swim larger distances and
access multiple cleaning stations while others access the only cleaning
station in their territory, it is crucial to categorize clients as
either ``visitors'' or ``residents'', respectively. During cleaning interactions,
a cleaner fish often face a choice between a visitor and a resident client
seeking its cleaning services simultaneously. Visitors have the option
to switch to another cleaner fish if being made to wait, while residents
must wait for inspection. Indeed,
visitors have been observed to use their partner choice option in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001}.

When researchers aimed at testing wild-caught cleaner fish, as
well as individuals from other species, abilities
to prioritize visitor over resident clients in a lab-based paradigm,
they used Plexiglas plates of different colours and shapes offering the same
amount of food as surrogates for visitor and resident clients.
One plate is made to behave like a resident, i.e.~it remains until
the cleaner fish has eaten off it. The other plate was made to behave
like a visitor, i.e.~it is removed if not inspected first.
This laboratory paradigm originally inspired by the cleaning market
of cleaner fish is more generally referred as the ephemeral reward task.
Individuals are assessed in their capacity to develop a preference for the
ephemeral option, which gives them access to two reward units.
Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Furthermore, adult cleaner fish outperformed various primates as well as rats and
pigeons in this original version of the ephemeral reward task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; yet the African grey
parrots solve this task as well \citep{pepperberg_Can_2014}.
This between variation in performance, and primates' known cognitive abilities
\(\color{red}{\text{ref}}\), shows that performance in the task
cannot be readily explained by classifying species as having high and
low cognitive abilities. Instead, this between species variation suggests
that solving the ephemeral reward task requires a set of cognitive
mechanisms that are not present in all the tested taxa, or were not used
under the experimental set-up.

Recent research on the cognitive abilities needed to solve the ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009}, who defined cognition as including
all ways in which animals acquire information
through the senses, process, retain and decide to act on it.
Indeed, perception of relevant cues is of major importance
for individual performance. For example, some animals find relevant
information in the plates \citep{wismer_Cuebased_2019} while others find
it in the food \citep{pretot_Comparative_2021, pretot_Comparing_2016}.
However, identifying cues as salient is not the only challenge
of the ephemeral reward task, as revealed by proximate learning models.
Such models allow varying the cognitive tool kit and evaluating
which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the ephemeral reward task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaner fish face not only resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well a resident alone or
a visitor alone. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible scenarios (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to distinguish and value
one stimulus differently alone from compound versions of it has been termed
configurational learning, chunking,
or segmentation (see references in \citep{prat_Modelling_2022}).

In addition to configurational learning, cleaners also need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Low temporal discounting is when
individuals include in their valuation of an action the reward effects
that this will have in the future. This is done by combining in a single
valuation the reward obtained in the current time with all the reward that
comes after, discounting for how far in the future reward
is accrued. `Chaining' the reward of these different time steps allows
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations \citep{enquist_Power_2016}. Even though, `chaining'
can be readily implemented computationally in learning models
\citep{enquist_Power_2016, sutton_Reinforcement_2018}, cognitively it seems to be
a complex adaptation \citep{suddendorf_Evolution_2007}. On the other hand,
using client behaviour as a negative reinforcer is, in principle,
easier to implement. Negative reinforcement is part of the ubiquitous
learning mechanism termed operant conditioning \citep{thorndike_Animal_1898, skinner_Behavior_1938}. Thus, the standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Ideally, however, the two
mechanisms should be evaluated in light of how well
they explain the available data. Note that different fields interested in
cognition and decision making use different words to refer to negative
reinforcers \citep{quinones_Reinforcement_2019, sutton_Reinforcement_2018}.
Here, for the sake of simplicity and clarity, we will use the word
`penalty' to refer to this mechanism which includes a negative reinforcer.

Over the last decade, our team tested over a hundred wild-caught
cleaner fish in the exact same paradigm of the ephemeral reward
task \citep{salwiczek_Adult_2012, wismer_Variation_2014, triki_Decrease_2018, triki_Biological_2019, triki_Brain_2020}. These fish often come from
different reef locations around the study field site at Lizard Island,
Great Barrier Reef in Australia. Further investigation of the local
eco-sociological conditions revealed that cleaner and client
fish population densities have a substantial impact on cleaner fish
performance in the task. Cleaner fish from reef sites with relatively
low densities were more likely to fail at solving the task \citep{triki_Biological_2019, triki_Decrease_2018, wismer_Variation_2014}. The explanation for this pattern
is twofold. First, low cleaner fish population densities imply
low supply and high demand for cleaning services.
Under such conditions, there are fewer occasions under
which failing to give priority to a visitor translates to an empty
cleaning station. Second and also related to the state of the cleaning market,
visitor clients at low cleaner fish population densities
are less likely to exert their partner choice options
and are hence more likely to wait for inspection if not given cleaning
priority \citep{triki_Brain_2020}. In the absence of partner choice
cleaner fish should not give priority to visitors anymore.

The interplay between the underlying cognitive machinery and
local ecological conditions apparently generate the documented variable
performance among individuals of the same species.
Our approach of fitting the computational model to the empirical data on
fish densities and cleaner fish censuses and performance
in the ephemeral reward task aimed at: (i), determining which
mechanism cleaner fish use to incorporate future consequences of current
decisions by testing whether chaining, penalty, or a combination
of both best explains their performance; (ii) determining whether the
two mechanisms differed with respect to the ecological conditions
that are likely to cause high versus low performance in the
ephemeral reward task. Additionally, we assessed which mechanism
yields optimal performance patterns. Relying on the logic of
biological market theory, we predicted
that appropriate performance is to show a low preference for visitors under low
local cleaner-to-client ratio.

@@

The model consists of a set of individual-based simulations where
individuals, representing cleaner fish, face a series of choices
between two options, which simulate the natural conditions
of the cleaning market. Individuals experience a series
of discrete time points in which they face different `states',
defined by the number and category of client fish (visitor or residents)
inviting for cleaning services. There are six possible states:
zero clients, one resident, one visitor,
resident-resident, visitor-visitor, and resident-visitor.
The probability of each state is largely determined by the relative
abundance of cleaner fish, residents and visitors, but to some degree by
cleaner fish choices when it faces the resident-visitor combination.
This is because residents are willing to queue for cleaning service;
while visitors leave the queue (with a certain probability) when made to wait.
Individuals obtain a fixed reward from cleaning a client fish
regardless of the category. Every time individuals face and make a choice
they update the probability of making that same choice. The update is
based on the difference between the expected reward and the obtained
reward (prediction error) \citep{sutton_Reinforcement_2018, rescorla_Theory_1972}. Given the new information, the update is carried
in the direction that leads to more reward being obtained, given
the new information. In the long run, the probability of choosing a visitor
over a resident converges in the model. To which probability the model
will converge depends on the relative abundance of cleaners, visitors
and residents; as well as on the probability of visitors leaving the
cleaning station when made to wait. Further details of the model
implementation can be found in Quiñones \emph{et al} \citep{quinones_Reinforcement_2019}.

The model shows that agents need to find a way
to incorporate future consequences of current choices. In the model,
this could be achieved with either of two parameters that could
also work together. First, \(\gamma\) measures how
much individuals include future rewards in their decision updates.
If \(\gamma=0\), individuals only use the immediate reward obtained from a
cleaning interaction. As \(\gamma\) increases, individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards of an action (chaining).
Second, \(\eta\) measures how much individuals include in their reward the
fleeing behaviour of visitors as a negative component (penalty). Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

@@

The cognitive performance data was from a total of 120 cleaners
(10 individuals per 12 site/year) tested in the
ephemeral reward task \citep{triki_Biological_2019, triki_Brain_2020}. Authors
housed all captured cleaners individually in glass aquaria
( \(62cm \times 27cm \times 37 cm\) ) and provided them
with PVC pipes (\(10 cm \times 1 cm\)) as shelters.
The task consisted of exposing the cleaner fish to substitute
models of client fish in the form of two \emph{Plexiglas} plates offering the
same amount of food (one item of mashed prawn). The two plates differed
in colour and pattern (horizontal green stripes or vertical pink stripes)
but had equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or
resident (permanent food source). That is, if a cleaner fish inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor plate
based solely on the behaviour-cue of the plates rather than size-cue
\citep{wismer_Cuebased_2019}. Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
tested the fish for a maximum of 200 trials with 20 trials a day, 10 trials
in the morning and 10 trials in the afternoon. They randomized and
counterbalanced the plates' spatial location (i.e.~left or right)
between trials. Similarly, they counterbalanced the plates' decoration
(colour and pattern) and the plates' role (visitor or resident) between the
tested fish. In the original studies, once a fish reach a learning criterion,
that is, performing significantly above chance level (\(> 50%
\), \(p-value ≥ 0.05\)),
they passed to a reversal version of the task where the roles of the
visitor/resident Plexiglas plates were swapped (see Triki et al.~2019, 2020).
Here, we used instead a subset of these data in order to have an idea of
cleaner fish preferences for the visitor plate, even if they do not reach
the learning criterion within 200 trials. To do so, we first extracted the
trial-by-trial outcomes from the last two sessions (20 trials) of those who
never reached the learning criterion for visitor plate
(\(N = 45\) cleaner fish).
For those who reached the learning criterion at some point during the
test and passed to a reversal phase, we extracted the trial-by-trial
outcomes from the last session (10 trials) before passing to reversal
and the last 10 trials they were exposed to in the test
(\(N = 75\) cleaner fish).
We chose a combination of initial and reversal to quantify preference
for the visitor client, given that it matches well the criteria chosen
in previous analysis of the ephemeral reward task experimental set-up
\citep{triki_Biological_2019, triki_Brain_2020}. In the
supplementary material (Fig. \ref{fig:rawdata}) we show how our choice maps
to the previously used criteria. For comparison, we also show a quantification
of visitor preference based only on the initial phase of the task.

@@

The aim of the analyses is to fit the key model parameters
\(\gamma\) and \(\eta\), to the empirical data from Triki \emph{et al.}
\citep{triki_Biological_2019, triki_Brain_2020} to test whether each or
a combination of these effects is a better explanation for the pattern
seen in the data. We used the ecological variables: cleaners, visitor clients,
resident clients abundances and visitor clients leaving probability,\\
as input to the models. As the response variable, we used cleaners'
preference for visitors over residents in the ephemeral reward task.
Finally, we used the preference for visitors resulting from the
model simulations as the prediction for the response variable.
We kept all other parameter values used for the model simulations constant,
see Table \ref{tab:param}.

To capture with the model the relationship between the ecological
variables and cleaner fish preferences for visitors, we needed to scale
the absolute population densities of cleaner fish from the empirical data
to a measure of relative abundances that captures client visitation patterns.
This is because, in the model, relative abundances of clients
define not only the probability of residents and visitors
but also how often the cleaning station is empty (e.g.~there are no
clients to be cleaned). The frequency with which clients visit the station
is another variable, which in nature may vary among different client species
depending on their ectoparasite loads. We do not have field estimates
for species-specific parasite loads, especially not as a function of the site.
In order to control for these aspects, we computed a measure of relative
cleaner fish abundance for each reef site relative to
absolute abundances and multiplied it
by a scaling constant that changes the range of the variable. This scaling
constant is hence meant to capture variation in the market conditions driven
partly by cleaner fish abundance. We fitted the value of the scaling constant as
part of the statistical inference. As for the visitor and resident
abundances, we computed a relative measure with respect to the
total client abundance, and weighted that by the rescaled relative absence
of cleaners (\(1-\text{relative cleaner abundance}\)). Thus, all three measures
of relative abundance sum up to 1, and can be used in the model as a proxy
for the probability of having different options in the cleaning station.
Note that we introduced the scaling constant to control for variation that
is not captured by the model; its parameter distribution does not offer
biological insights.
